\documentclass{beamer}

\usecolortheme{beaver}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{itemize items}[square]
\setbeamercolor{block title}{bg=gray!20, fg=red!70!black}

\usepackage{booktabs}
\usepackage{fancyvrb}
\usepackage{fvextra}
\usepackage{hyperref}
\usepackage{tabularx}

\DefineVerbatimEnvironment{CodeBlock}{Verbatim}{
    fontsize=\tiny,
    frame=single,
    rulecolor=\color{gray},
    framesep=2mm,
    breaklines=true,
    breakanywhere=true,
    breaksymbolleft={},
    breakautoindent=true,
    breakindent=1em
}

\title{RKT/ICK - INTERCAL embedded in RACKET}
\subtitle{From INTERCAL Source to Racket State Machines}
\author{Scheming Racketeers}
\date{April 27th, 2026}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \frametitle{Project Goal}

  \vfill
  \begin{itemize}
    \item Implement a substantial, source-compatible subset of C-INTERCAL in Racket.
    \item Support real programs, not just toy syntax fragments.
    \item Compile INTERCAL into ordinary Racket by macro expansion instead of interpreting an external bytecode.
    \item Preserve difficult semantics: \texttt{COME FROM}, \texttt{NEXT}/\texttt{RESUME}, \texttt{ABSTAIN}, gerunds, and tape I/O.
  \end{itemize}
  \vfill
\end{frame}

\begin{frame}
  \frametitle{What the Project Implements}

  \scriptsize
  \begin{tabularx}{\textwidth}{@{}l>{\raggedright\arraybackslash}X@{}}
    \toprule
    Area & Implemented functionality \\
    \midrule
    Reader/frontend & Source cleaning, continuation-line merging, lexer, brag grammar, parse-tree normalization \\
    Core statements & Assignment, \texttt{NEXT}, \texttt{COME FROM}, \texttt{READ OUT}, \texttt{WRITE IN}, \texttt{STASH}, \texttt{RETRIEVE}, \texttt{IGNORE}, \texttt{REMEMBER}, \texttt{FORGET}, \texttt{RESUME}, \texttt{GIVE UP}, \texttt{TRY AGAIN}, \texttt{NOTHING} \\
    Control-state modifiers & \texttt{NOT}, \texttt{ONCE}, \texttt{AGAIN}, percentage/chance execution, politeness checking \\
    Dynamic control & Label abstention, gerund abstention/reinstatement, delayed \texttt{COME FROM}, counted abstain state \\
    Data model & Onespots, twospots, tails/hybrids with multidimensional arrays, packed \texttt{SUB}, \texttt{BY}-dimensioned arrays \\
    Operators & \texttt{MINGLE}, \texttt{SELECT}, unary AND/OR/XOR, width-aware rewrites, constant folding \\
    I/O & Numeric input/output plus C-INTERCAL-style array tape I/O \\
    Libraries & \texttt{syslib.i} integration and label-triggered library loading \\
    \bottomrule
  \end{tabularx}
\end{frame}

\begin{frame}[fragile]
  \frametitle{INTERCAL Control-Flow Semantics}

  \begin{itemize}
    \item \texttt{NEXT} pushes a return target; \texttt{RESUME} removes one or more entries and jumps to the last removed target.
    \item \texttt{COME FROM} is resolved after the triggering line executes, with delayed behavior for resumed \texttt{NEXT} entries.
    \item \texttt{ABSTAIN} and \texttt{REINSTATE} are modeled with counted runtime state so overlapping abstentions compose correctly.
    \item \texttt{ONCE} and \texttt{AGAIN} mutate line state after execution or skip.
  \end{itemize}

  \vspace{0.2cm}
  \begin{block}{Why this matters}
    The hard part of INTERCAL is not parsing the keywords. It is compiling bizarre, self-modifying control flow into something predictable enough for Racket to run.\footnote{This is still an open area of improvement, as the performance of our implementation could be greatly improved.}
  \end{block}
\end{frame}



\begin{frame}[fragile]
  \frametitle{Pipeline Overview}

  \begin{enumerate}
    \item Clean source text and merge continuation lines.
    \item Tokenize and parse INTERCAL into a concrete syntax tree.
    \item Normalize that tree into a strict S-expression IR.
    \item Reader emits a Racket module that calls the backend macro layer.
    \item Macros add library code, lower the IR into line tuples, analyze it, and emit explicit Racket state machines.
    \item Runtime helpers enforce INTERCAL semantics and produce output.
  \end{enumerate}

  \vspace{0.2cm}
  \textbf{Representative source program}

  \input{snippets/sample-source.tex}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Clean Source and Merge Continuation Lines}

  \begin{columns}[T]
    \begin{column}{0.62\textwidth}
      \textbf{Cleaner}

      \input{snippets/cleaner-snippet.tex}
    \end{column}
    \begin{column}{0.33\textwidth}
      \begin{itemize}
        \item The cleaner keeps only lines that look like INTERCAL statements or valid continuations.
        \item Indented continuation lines are merged into the current logical line.
        \item Non-statement prose gets dropped before the grammar ever sees it.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Cleaning Example}

  \begin{columns}[T]
    \begin{column}{0.48\textwidth}
      \textbf{Raw input}
      \input{snippets/sample-raw-source.tex}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{Cleaned logical lines}
      \input{snippets/sample-cleaned-source.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=5]
  \frametitle{Tokenize}

  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Lexer}

      \input{snippets/lexer-snippet.tex}
    \end{column}
    \begin{column}{0.43\textwidth}
      \begin{itemize}
        \item The lexer turns cleaned source text into tokens the grammar can consume.
        \item It also expands packed \texttt{SUB} forms before parsing so the grammar sees a regularized token stream.
        \item Token kinds like numbers, assignment markers, labels, and keywords are made explicit here.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Tokenization Example}

  \begin{columns}[T]
    \begin{column}{0.42\textwidth}
      \textbf{Cleaned source line}
      \input{snippets/sample-line-source.tex}
    \end{column}
    \begin{column}{0.54\textwidth}
      \textbf{Token stream}
      \input{snippets/sample-line-tokens.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=5]
  \frametitle{Parse Tokens into a Concrete Syntax Tree}

  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Grammar Example}

      \input{snippets/grammar-snippet.tex}
    \end{column}
    \begin{column}{0.43\textwidth}
      \begin{itemize}
        \item The grammar, written with brag, parses those tokens into a concrete syntax tree.
        \item Arrays and dimensions are parsed uniformly so they can normalize into one IR.
        \item The tree still reflects source-level structure like labels, prefixes, postfix modifiers, and operator nesting.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Parsing Example}

  \begin{columns}[T]
    \begin{column}{0.37\textwidth}
      \textbf{Token Stream}
      \input{snippets/sample-line-tokens.tex}
    \end{column}
    \begin{column}{0.59\textwidth}
      \textbf{Partial Concrete syntax tree}
      \input{snippets/sample-line-parse.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=15]
  \frametitle{Normalize the CST into IR}

  \begin{columns}[T]
    \begin{column}{0.56\textwidth}
      \textbf{Normalizer Excerpt}

      \input{snippets/normalizer-plain-snippet.tex}
    \end{column}
    \begin{column}{0.39\textwidth}
      \begin{itemize}
        \item Every program has one top-level form: \texttt{sick-program/syslib}.
        \item Labeled lines become \texttt{(<label> <statement>)} pairs.
        \item Grammar scaffolding, punctuation, and keyword strings disappear.
        \item Variables, constants, labels, array subscripts, and dimensions all normalize into fixed arities the macros can match directly.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=5]
  \frametitle{Normalization Example}

  \begin{columns}[T]
    \begin{column}{0.48\textwidth}
      \textbf{Concrete parse tree}
      \input{snippets/sample-parse.tex}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{Strict S-expression IR}
      \input{snippets/sample-normalized.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Reader Code}

  \begin{columns}[T]
    \begin{column}{0.62\textwidth}
      \textbf{Reader}
      \input{snippets/reader-snippet.tex}
    \end{column}
    \begin{column}{0.33\textwidth}
      \begin{itemize}
        \item The reader wraps the normalized program in an ordinary Racket module.
        \item The module exports \texttt{intercal-main}; execution enters through the backend macro form, not an interpreter loop.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Reader Output Example}

  \begin{columns}[T]
    \begin{column}{0.47\textwidth}
      \textbf{Normalized IR}
      \input{snippets/sample-normalized.tex}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{Reader output}
      \input{snippets/sample-module.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=15]
  \frametitle{Macro-Lower IR into Racket State Machines}

  \begin{columns}[T]
    \begin{column}{0.57\textwidth}
      \input{snippets/macro-snippet.tex}
    \end{column}
    \begin{column}{0.38\textwidth}
      \begin{itemize}
        \item The reader emits \texttt{sick-program/syslib} around the normalized user IR.
        \item \texttt{sick-program/syslib} adds needed library code and lowers the combined IR into \texttt{sick-program-core} line tuples.
        \item \texttt{sick-program-core} analyzes variables, abstain targets, ignored variables, \texttt{COME FROM} targets, and gerunds.
        \item Expansion emits optimized Racket state bindings and dispatch clauses.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=12]
  \frametitle{Macro Expansion Example}

  \begin{columns}[T]
    \begin{column}{0.48\textwidth}
      \textbf{Lowered \texttt{sick-program-core} input}
      \input{snippets/sample-macro-input.tex}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{Expanded Racket shape}
      \input{snippets/sample-state-machine-shape.tex}
    \end{column}
  \end{columns}

  \vspace{0.15cm}
  \scriptsize
  The extra fields are compile-time metadata: runtime line number, source label,
  politeness, execution percentage, \texttt{NOT}, \texttt{ONCE}, \texttt{AGAIN},
  and the normalized operation.
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Runtime Helpers}

  \begin{columns}[T]
    \begin{column}{0.58\textwidth}
      \input{snippets/runtime-snippet.tex}
    \end{column}
    \begin{column}{0.37\textwidth}
      \begin{itemize}
        \item The expanded state machine closes over runtime state such as \texttt{output-acc}, \texttt{next-stack}, abstain counts, and ignore flags.
        \item Helpers enforce INTERCAL-sized storage, numeric input, tape-style array I/O, and INTERCAL error reporting.
        \item \texttt{get-actual-next} applies delayed \texttt{COME FROM} control after a line executes.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}[fragile,shrink=8]
  \frametitle{Runtime Execution Example}

  \begin{columns}[T]
    \begin{column}{0.48\textwidth}
      \textbf{Program entering runtime}
      \input{snippets/sample-runtime-source.tex}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{Observed runtime behavior}
      \input{snippets/sample-runtime-result.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Missing or Incomplete Features}

  \vfill
  \begin{itemize}
    \item \texttt{CREATE} is not implemented.\footnote{This feature is a known painpoint, and is disabled by default in C-INTERCAL.}
    \item Runtime Syntax errors.\footnote{Given our compliation pipeline, it is not trivial for us to implement this feature of INTERCAL.}
    \item Multithreading/backtracking features are not implemented.
    \item PIC-INTERCAL specific functionality is not implemented.
    \item TriINTERCAL and operand-overloading extensions are not implemented.
    \item OIL optimizer idiom ingestion is not implemented; only built-in conservative optimizations exist.
    \item Some upstream pit programs still diverge or misbehave, so compatibility is substantial but not complete.
  \end{itemize}
  \vfill
\end{frame}

\begin{frame}[fragile]
  \frametitle{Evidence: Real Programs}

  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Programs currently exercised in the repository}
      \begin{itemize}
        \item \texttt{pit/fib.i}
        \item \texttt{pit/hanoi.i}
        \item \texttt{pit/onceagain.i}
        \item \texttt{pit/hello.i}
        \item \texttt{pit/ignorret.i}
        \item \texttt{pit/beer.i}
        \item \texttt{pit/life.i}
        \item \texttt{pit/rot13.i}
        \item \texttt{pit/unlambda.i}
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Example program}
      \input{snippets/hello-snippet.tex}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{User-Facing Documentation}

  \begin{itemize}
    \item We added Scribble documentation for language usage, supported features, semantic gaps, and the compilation pipeline.
    \item The documentation covers both INTERCAL and our S-expression variant, with any semantic information instead being deferred to Eric Raymonds docs.
    \item The same documentation is also avaible from Racket's official docs website.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Takeaways}

  \begin{itemize}
    \item The project implements a straigtforward compiler pipeline: source text $\rightarrow$ CST $\rightarrow$ normalized S-expr IR $\rightarrow$ macro-expanded Racket.
    \item The implementation is now capable of running nontrivial upstream programs, including a language implementation built on top of another esolang implementation.
    \item The remaining items, should they be done, are related to completeness: extensions, more example program coverage, and packaging polish.
  \end{itemize}
\end{frame}

\end{document}
